Higher-Order unification for free!

### 2024-07-26 paper/paper_for_taps/source_src/code/maybe_eta.tex
new version (seems accepted by taps)

\begin{elpicode}
  kind arity type.
  type arity nat -> arity.
  kind ovariable type.
  type ov addr -> ovariable.
  kind mvariable type.
  type mv addr -> arity -> mvariable.
  kind mapping type.
  ~\PYG{k+kd}{type} \PYG{n+nf}{(<->)} \PYG{k+kt}{ovariable -> mvariable -> mapping}~.
  typeabbrev mmap (list mapping).

  typeabbrev scope (list tm).
  typeabbrev inctx ho.inctx.
  kind baselink type.
  type link-eta  tm -> tm -> baselink.
  type link-llam tm -> tm -> baselink.
  typeabbrev links (list (inctx baselink)).
  typeabbrev link (inctx baselink).

  macro @val-link-eta T1 T2 :- ho.val (link-eta T1 T2).
  macro @val-link-llam T1 T2 :- ho.val (link-llam T1 T2).

  type get-lhs link -> tm -> o.
  get-lhs (val (link-llam A _)) A.
  get-lhs (val (link-eta A _)) A.

  type get-rhs link -> tm -> o.
  get-rhs (val (link-llam _ A)) A.
  get-rhs (val (link-eta _ A)) A.

  type occurs-rigidly to -> to -> o.
  occurs-rigidly N N.
  occurs-rigidly _ (o-app [o-uva _|_]) :- !, fail.
  occurs-rigidly N (o-app L) :- exists (occurs-rigidly N) L.
  occurs-rigidly N (o-lam B) :- pi x\ occurs-rigidly N (B x).

@@

  type maybe-eta to -> list to -> o.                  ~\customlabel{rule:maybeeta}{(\maybeeta)}~
  maybe-eta (o-app[o-uva _|Args]) L :- !,
    forall1 (x\ exists (reducible-to [] x) Args) L, !. 
  maybe-eta (o-lam B) L :- !, pi x\ maybe-eta (B x) [x | L].
  maybe-eta (o-app [T|Args]) L :- (name T; T = o-con _),
    split-last-n {len L} Args First Last,
    none (x\ exists (y\ occurs-rigidly x y) First) L,
    forall2 (reducible-to []) {rev L} Last.

  type locally-bound tm -> o.
  type get-scope-aux tm -> list tm -> o.
  get-scope-aux (con _) [].
  get-scope-aux (uva _ L) L1 :- 
    forall2 get-scope-aux L R,
    flatten R L1.
  get-scope-aux (lam B) L1 :- 
    pi x\ locally-bound x => get-scope-aux (B x) L1.
  get-scope-aux (app L) L1 :- 
    forall2 get-scope-aux L R,
    flatten R L1.
  get-scope-aux X [X] :- name X, not (locally-bound X).
  get-scope-aux X [] :- name X, (locally-bound X).

  type names1 list tm -> o.
  names1 L :-
    names L1, 
    new_int N,
    if (1 is N mod 2) (L1 = L) (rev L1 L).

  type get-scope tm -> list tm -> o.
  get-scope T Scope :-
    get-scope-aux T ScopeDuplicata,
    undup ScopeDuplicata Scope.
  type rigid to -> o.
  rigid X :- not (X = o-uva _).

  type comp-lam (to -> to) -> (tm -> tm) -> 
    mmap -> mmap -> links -> links -> subst -> subst -> o.
  comp-lam F G M1 M2 L1 L3 S1 S2 :-
    pi x y\ (pi M L S\ comp x y M M L L S S) =>         ~\customlabel{rule:complamh}{(\ensuremath{H_\lambda})}~
      comp (F x) (G y) M1 M2 L1 (L2 y) S1 S2,
    close-links L2 L3.

  type close-links (tm -> links) -> links -> o.
  close-links (v\[X v|L v]) [abs X|R] :- close-links L R.
  close-links (_\[]) [].
  type comp to -> tm -> mmap -> mmap -> links -> links ->
    subst -> subst -> o.
  comp (o-con C) (con C) M M L L S S.
  comp (o-lam F) (uva A Scope) M1 M2 L1 L3 S1 S3 :-
    maybe-eta (o-lam F) [], !,
      alloc S1 A S2,
      comp-lam F F1 M1 M2 L1 L2 S2 S3,
      get-scope (lam F1) Scope, 
      L3 = [val (link-eta (uva A Scope) (lam F1)) | L2].
  comp (o-lam F) (lam F1) M1 M2 L1 L2 S1 S2 :-            ~\customlabel{rule:complam}{(\ensuremath{c_\lambda})}~
    comp-lam F F1 M1 M2 L1 L2 S1 S2.
  comp (o-uva A) (uva B []) M1 M2 L L S1 S2 :- 
    m-alloc (ov A) (mv B (arity z)) M1 M2 S1 S2.
  comp (o-app [o-uva A|Ag]) (uva B Ag1) M1 M2 L L S1 S2 :-
    pattern-fragment Ag, !,
      fold6 comp Ag Ag1 M1 M1 L L S1 S1,
      len Ag Arity, 
      m-alloc (ov A) (mv B (arity Arity)) M1 M2 S1 S2.
  comp (o-app [o-uva A|Ag]) (uva B Sc) M1 M3 L1 L3 S1 S4 :- !,
    pattern-fragment-prefix Ag Pf Extra, alloc S1 B S2,
    len Pf Ar, m-alloc (ov A) (mv C (arity Ar)) M1 M2 S2 S3,
    fold6 comp Pf    Pf1    M2 M2 L1 L1 S3 S3,
    fold6 comp Extra Extra1 M2 M3 L1 L2 S3 S4,
    Beta = app [uva C Pf1 | Extra1],
    get-scope Beta Sc, 
    L3 = [val (link-llam (uva B Sc) Beta) | L2].
  comp (o-app A) (app A1) M1 M2 L1 L2 S1 S2 :-            ~\customlabel{rule:compapp}{(\ensuremath{c_@})}~
    fold6 comp A A1 M1 M2 L1 L2 S1 S2.

  type alloc mem A -> addr -> mem A -> o.
  alloc S N S1 :- mem.new S N S1.

  type compile-terms-diagnostic 
    triple diagnostic to to -> 
    triple diagnostic tm tm -> 
    mmap -> mmap -> 
    links -> links -> 
    subst -> subst -> o.
  compile-terms-diagnostic (triple D FO1 FO2) (triple D HO1 HO2) M1 M3 L1 L3 S1 S3 :-
    beta-normal FO1 FO1',
    beta-normal FO2 FO2',
    comp FO1' HO1 M1 M2 L1 L2 S1 S2,
    comp FO2' HO2 M2 M3 L2 L3 S2 S3.

  type compile-terms 
    list (triple diagnostic to to) -> 
    list (triple diagnostic tm tm) -> 
    mmap -> links -> subst -> o.
  compile-terms T H M L S :- 
    fold6 compile-terms-diagnostic T H [] M_ [] L_ [] S_,
    print-compil-result T H L_ M_, 
    deduplicate-map M_ M S_ S L_ L.

  type make-eta-link-aux nat -> addr -> addr -> 
    list tm -> links -> subst -> subst -> o.
  make-eta-link-aux z Ad1 Ad2 Scope1 L H1 H1  :-
    rev Scope1 Scope, eta-expand (uva Ad2 Scope) T1,
    L = [val (link-eta (uva Ad1 Scope) T1)].
  make-eta-link-aux (s N) Ad1 Ad2 Scope1 L H1 H3 :-
    rev Scope1 Scope, alloc H1 Ad H2, 
    eta-expand (uva Ad Scope) T2,
    (pi x\ make-eta-link-aux N Ad Ad2 [x|Scope1] (L1 x) H2 H3),
    close-links L1 L2,
    L = [val (link-eta (uva Ad1 Scope) T2) | L2].

  type make-eta-link nat -> nat -> addr -> addr -> 
          list tm -> links -> subst -> subst -> o.
  make-eta-link (s N) z Ad1 Ad2 Vars L H H1 :- 
    make-eta-link-aux N Ad2 Ad1 Vars L H H1.
  make-eta-link z (s N) Ad1 Ad2 Vars L H H1 :- 
    make-eta-link-aux N Ad1 Ad2 Vars L H H1.
  make-eta-link (s N) (s M) Ad1 Ad2 Vars Links H H1 :-
    (pi x\ make-eta-link N M Ad1 Ad2 [x|Vars] (L x) H H1),
    close-links L Links.

  type deduplicate-map mmap -> mmap -> 
      subst -> subst -> links -> links -> o.
  deduplicate-map [] [] H H L L.
  deduplicate-map [((ov O <-> mv M (arity LenM)) as X1) | Map1] Map2 H1 H3 L1 L3 :-
    take-list Map1 ((ov O <-> mv M' (arity LenM'))) _, !,
    std.assert! (not (LenM = LenM')) "Deduplicate map, there is a bug",
    print "arity-fix links:" {ppmapping X1} "~!~" {ppmapping ((ov O <-> mv M' (arity LenM')))}, 
    make-eta-link LenM LenM' M M' [] New H1 H2,
    print "new eta link" {pplinks New},
    append New L1 L2,
    deduplicate-map Map1 Map2 H2 H3 L2 L3.
  deduplicate-map [A|As] [A|Bs] H1 H2 L1 L2 :- 
    deduplicate-map As Bs H1 H2 L1 L2, !.
  deduplicate-map [A|_] _ H _ _ _ :- 
    halt "deduplicating mapping error" {ppmapping A} {ho.ppsubst H}.
\end{elpicode}

>>>>

\begin{elpicode}
  type occurs-rigidly to -> to -> o.
  occurs-rigidly N N.
  occurs-rigidly _ (o-app [o-uva _|_]) :- !, fail.
  occurs-rigidly N (o-app L) :- exists (occurs-rigidly N) L.
  occurs-rigidly N (o-lam B) :- pi x\ occurs-rigidly N (B x).

@@

  type maybe-eta to -> list to -> o.                  ~\customlabel{rule:maybeeta}{(\maybeeta)}~
  maybe-eta (o-app[o-uva _|Args]) L :- !,
    forall1 (x\ exists (reducible-to [] x) Args) L, !. 
  maybe-eta (o-lam B) L :- !, pi x\ maybe-eta (B x) [x | L].
  maybe-eta (o-app [T|Args]) L :- (name T; T = o-con _),
    split-last-n {len L} Args First Last,
    none (x\ exists (y\ occurs-rigidly x y) First) L,
    forall2 (reducible-to []) {rev L} Last.
\end{elpicode}
